%\acknowledgments[6]{
\startacknowledgementspage

I became a graduate student in the University of Michigan in the fall of 2008. Five years have passed and I would like to thank many people during this long and unique Ph.D. journey in my life. The completion of this dissertation would not be possible without anyone of them.

First of all, I would like to sincerely thank my advisor, Professor Z. Morley Mao. I first met her when I was an intern in Microsoft Research Asia in 2007, wondering about my future, and she offered me much helpful advice for pursuing a Ph.D. degree in the U.S. Morley has provided me excellent and professional guidance on various research projects throughout my Ph.D. study, given her expertise in the computer networking and mobile computing. Moreover, she helped me gain the confidence, ability and most importantly, the enthusiasm to tackle difficult real-world problems. Her diligence and enthusiasm has set a great model for me and the experiences working with her would be my life-long treasure.

I am greatly thankful for my mentors Kun Tan and Yongguang Zhang in Microsoft Research Asia (Beijing 2007), Yinglian Xie and Fang Yu in Microsoft Research Silicon Valley (California 2010), Subhabrata Sen and Oliver Spatscheck in AT\&T Labs - Research (New Jersey 2012). I am fortunate to work with these knowledgeable and excellent researchers. I enjoyed my internships, which would be my unforgotten memories. I also enjoyed the conversations with them about exciting research ideas, and about career and life as well. Many of their words have kept ringing in my mind, helping me to improve.

I would like to express my deep gratitude to Professor Jason Flinn, Professor Robert Dick, and Dr. Subhabrata Sen for serving as my thesis committees. Their valuable suggestions and comments have greatly helped me to improve this dissertation. My grateful thanks are also due to Professor Scott Mahlke, Professor J. Alex Halderman for their excellent teaching on my core computer science courses, and Dr. Ming Zhang and Dr. Paramvir Bahl for their priceless help for me in writing and presenting the first paper~\cite{mobisys.3gtest} in my life.

It is hard for me to forget the great professors during my undergraduate study in Tsinghua University who taught me the basics in computer science and helped me find the joy in this area. Their names include Xiping Mao, Wei Chen, Hong Wang, Jianhua Feng, Zhongxin Xu (who has passed away in 2010 and is long remembered), Shimin Hu, Ning Su, \etc I owe my special thanks to Professor Andrew Chi-Chih Yao, who included me in his first special pivot class, where I have received great education in computer science, and he himself also demonstrated what a great computer scientist and a Turing Award winner is like.

I would like to thank all my friends and colleagues at Michigan. I have worked closely with Dr Feng Qian and Qiang Xu since the first day of my Ph.D. study. Dr. Ying Zhang, Dr. Xu Chen and Dr. Zhiyun Qian are former members of our research group and they are also my good friends. I am grateful for my great roommate Lujun Fang and his wife Yiwen Mu. My thanks also go to my other friends and colleagues in the Computer Science department at Michigan, including, but not limited to Bin Liu, Yunjing Xu, Yudong Gao, Birjodh Tiwana, Zhaoguang Wang, Lide Zhang, Mark Gordon, Caoxie Zhang, Fangjian Jin, Zhe Chen, Kee Shen Quah, Li Qian, Sanae Rosen, Bangxin Hu, Jie Yu, Xinyu Zhang, Timur Alperovich, Yihua Guo, Yuanyuan Zhou, Haokun Luo, Qi Chen. I am thankful for University of Michigan Table Tennis Club (UMTTC) and University of Michigan Chinese Drama Club for brining to me many good times and many good friends: Hang Zhao, Weilun Xu, Allen Hung, Ason Chiang, Bo Bi, Yang Zhou, Tomas Fuentes-Afflick; Lang Ming, Yuhua Wang, Yueyang Zhong, Siyi Fang, Beimin Zhu, \etc I want to thank my piano teachers Alexandra Lynelle James, Yuanchang Zhou and Shuai Wang, my table tennis coaches David Kleshchik, Naihui Liu and Qing Miao. I would also like to thank my other friends outside Michigan for appearing in and shaping my life: Wentao Han, Jianfeng Gao, Jingyue Wu, Hui Min, Ruizhe Li, Chengwei Guo, Yifang Yuan, Jiajie Zhu, Haohui Mai, Yini Xu, Ning Ding, Shan Zhou, Zhaojie Cheng, Dongdong Peng, Simin Zhang, Tian Guo, Yi Chen, Zhenqiang Gong.

I could not love the University of Michigan more. ``For today, Goodbye; For tomorrow, Good Luck; Forever, GO BLUE!'' I will remember Ann Arbor forever, the most beautiful town in the world.

Finally, I would like to express my deepest thanks to my beloved parents, Xuemeng Huang and Zhaoxia Wang, and my dearest grandparents Shu Huang, Lingzhi Liu, Shengqin Wang (who passed away in 2008 to my greatest sadness) and Xiuhua Ma for their enduring love, support and encouragement throughout my life. This dissertation is dedicated to them.

\label{ACKNOWLEDGEMENTS}
